---
title: "HW5 - Make Data Move"
author: ""
date: "05/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore global development with R
Today, you will load a filtered gapminder dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks below.**

## Get the necessary packages
First, start with installing the relevant packages 'tidyverse', 'gganimate', and 'gapminder'.

```{r libraries, echo = FALSE}
install.packages("gganimate")
install.packages("gifski")
install.packages("av")
install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)
```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
head(gapminder)
```

The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```

...

We see an interesting spread with an outlier to the right. Answer the following questions, please:


1. *Why does it make sense to have a log10 scale on x axis?*
 It makes sense since the numbers are too big to have on the axis. The log10() is then smart since it works from a base 10-system, where numbers are reduced to a base instead of being whole numbers.
2. *Who is the outlier (the richest country in 1952 - far right on x axis)?* 
The richest country was Kuwait which i know since iøve used the "filter"-function (1952), based un the "country" and "gdpPercap"-column (selects-methods). Lastly i use the "arrange"-function to get the countries in a descending order (desc-method) based on the gdpPercap.
```{r richest country in 1952}
gapminder %>% 
  filter(year == 1952) %>% 
  select(country, gdpPercap) %>% 
  arrange(desc(gdpPercap))
```

Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

Tasks:

3. *Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation", which you might want to eliminate)*
```{r new 2007}
options(scipen = 999) #Remove the scientific notation
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10()+
    labs(y= "life expectancy", x = "GDP per capita") #I change the name of axis-titles using the "labs"-function (labs = labels).
#i change the colors based on continent using the "color"-method.
```


4. *What are the five richest countries in the world in 2007?*
Same code as in 1952, but this time i've used the "head"-method in order of finding the top-5.

```{r richest country in 2007}
gapminder %>% 
  filter(year == 2007) %>% 
  select(country, gdpPercap) %>% 
  arrange(desc(gdpPercap)) %>% 
  head(5)

```



##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10()  # convert x to log scale
anim
```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)

```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smoothes the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim2
```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 


5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)*
```{r anim3}
anim3 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  labs(title = 'Year: {frame_time}') +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim3
```

6. *Can you made the axes' labels and units more readable? Consider expanding the abreviated lables as well as the scientific notation in the legend and x axis to whole numbers.*
I use the "labs"-function to make make the title, which I call "Year:" and then use the inbuilt function "{frame_time}" to get the time from the dataset that fits to each frame.
I use the "theme"-function where i target each of the things i want ot enlarge. The title, axis' etc. all have a name in R which i use in the "plot"-function and then i pass the parameter "size" and the given fontsize i want it to be. Also, i changed the face of the title to become bold.
I dont get the last part of the assignment-description since the numbers aldready are whole numbers?
```{r anim4}
anim4 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  labs(title = 'Year: {frame_time}') +
  theme(axis.text=element_text(size=16), axis.title=element_text(size=20), plot.title=element_text(size=30, face="bold"), legend.text=element_text(size=16), legend.title=element_text(size=20)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim4
```

7. *Come up with a question you want to answer using the gapminder data and write it down. Then, create a data visualisation that answers the question and explain how your visualization answers the question. (Example: you wish to see what was mean life expectancy across the continents in the year you were born versus your parents' birth years). [Hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset and download more at https://www.gapminder.org/data/ ]*


